Open archaeology, open source?

### 2024-03-21 analysis/_01-introduction.qmd
Work more references into introduction

In their seminal vision for open archaeology, Beck and Neylon [-@beck2012 480-481] identified the movement as comprising a series of principles and practices "predicated on promoting open redistribution and access to the data, processes and syntheses generated within the archaeological domain" with the aim of "maximizing transparency, reuse and engagement while maintaining professional probity". Open archaeology therefore promotes more thoughtful scholarly communication practices, most notably in operations relating to publishing, data-sharing, education, and review processes. Archaeologists are also actively engaged in open source software development as means of sharing their research processes and creating tools and resources for general use.

The relationship between open science and open source is complicated by rhetorical claims that have a questionable connection to how academics actually _do_ open source. 
Does academic open source actually make research processes more transparent and improve research outcomes? 
Is it actually boosting efficiency by establishing a common store of knowledge and productive code? 
Is it actually helping to foster new globe-spanning connections and lead to novel research trajectories that would not otherwise come to pass?
Basically, is there more to it than just uploading code and data files to the internet?

Despite hopeful aspirations espoused by open archaeology advocates [see @kansa2014; @kintigh2015], these outcomes are not a given. As per @nguyen2022, @pownall2023 and @leonelli2023 we believe that these outcomes only actually arise in contexts where participants adhere to and are motivated by warrants, professional norms, and governance strategies that encourage these results. However, practical circumstances and systemic value regimes that frame what it means to work as an archaeologist presently inhibit the potential for radical transformation, even among open science's most ardent supporters.

>>>>

A 2012 special issue of *World Archaeology* marked the coming of age of 'open archaeology',
a new, digitally-oriented archaeology "predicated on promoting open redistribution and access to the data, processes and syntheses generated within the archaeological domain" with the aim of "maximizing transparency, reuse and engagement while maintaining professional probity" [@beck2012 480-481],
situated within the wider the open science movement [@lake2012 and @beck2012].
In the same issue, @ducke2012 specifically called attention to software, calling on archaeologists to more actively engage in open source practices – making software with source code available for anyone to freely access, modify or reuse.
Open source, though originating outside academy, emerged as an important component and complement of open science, sharing its vision for a communal, self-correcting and transparent mode of knowledge production.
Where before archaeologists had relied almost exclusively on "hand-me-down" software re-purposed from other fields [@scollar1999], open archaeology envisaged a community of archaeological research software engineers oriented around open source development practices and tools.

More than decade on, we are in a position to ask whether this hopeful early rhetoric of open archaeology has been borne out in actually-existing open source research software engineering in the field. 
Does academic open source actually make research processes more transparent and improve research outcomes? 
Is it actually boosting efficiency by establishing a common store of knowledge and productive code? 
Is it actually helping to foster new globe-spanning connections and lead to novel research trajectories that would not otherwise come to pass?
In other words, is there more to 'open archaeology' than just uploading text files to the internet?

The aspirations of open archaeology's early proponents [e.g. @kansa2014; @kintigh2015] were tempered by notes of caution that "the reward structures in academic and professional archaeology do little to incentivise participation in Open Archaeology" [@lake2012, echoing @beck2012, @kansa2012, @huggett2012 and @limp2011].
Following recent work by @nguyen2022, @pownall2023 and @leonelli2023, we believe that the outcomes listed above indeed only arise in contexts where there are organization structures, governance strategies, and professional norms that encourage them.
Thus practical circumstances and systemic value regimes that frame what it means to work as an archaeologist presently inhibit the potential for radical transformation, even among open science's most ardent supporters.
